---
title: "The Quiet Decline of Mental Health"
author: "Ruchi"
format:
  html:
    embed-resources: true
---

## The question

People carry real mental distress and still show up to work every day. We read "still functioning" as "fine." I wanted to see what the national numbers actually say.

## What the data shows

```{r}
#| echo: false
#| message: false
#| warning: false
library(cansim)
library(tidyverse)

mh <- get_cansim("13-10-0096")

canada_trend <- mh %>%
  filter(Indicators == "Perceived mental health, fair or poor",
         GEO == "Canada (excluding territories)",
         Sex == "Both sexes",
         `Age group` == "Total, 12 years and over",
         Characteristics == "Percent")

ggplot(canada_trend, aes(x = REF_DATE, y = VALUE, group = 1)) +
  geom_line(linewidth = 1.2, color = "#b22222") +
  geom_point(size = 2.5, color = "#b22222") +
  labs(
    title = "More Canadians are not okay",
    subtitle = "Share rating their mental health as fair or poor, 2015-2022",
    x = NULL,
    y = "Percent"
  ) +
  theme_minimal()
```

The share of Canadians rating their mental health as fair or poor more than doubled between 2015 and 2022, climbing every year and accelerating toward the end. This happened while the country kept working.

## What this shows and what it doesn't

This is the national picture. It does not yet break the trend down by who is employed, or by occupation, so it points to the pattern without isolating working people specifically.

## Work Absence in Health Care: A Structural Pattern, Not Just a Pandemic Spike

Using Statistics Canada's Work Absence Rates data (Table 14-10-0191), I examined 
how many workdays health care and social assistance workers lose per year compared 
to other sectors, and what's actually driving that loss.

```{r}
#| echo: false
#| message: false
#| warning: false

absence_data <- get_cansim("14-10-0191")
absence_data$REF_DATE <- as.numeric(absence_data$REF_DATE)

healthcare_absence <- absence_data %>%
  filter(`North American Industry Classification System (NAICS)` == "Health care and social assistance",
         REF_DATE >= 2015)

compare_data <- absence_data %>%
  filter(`Work absence statistics` == "Total days lost per worker in a year",
         Gender == "Total - Gender",
         `North American Industry Classification System (NAICS)` %in% 
           c("Health care and social assistance", "Service-producing industries"),
         REF_DATE >= 2015)

healthcare_split <- absence_data %>%
  filter(`North American Industry Classification System (NAICS)` == "Health care and social assistance",
         Gender == "Total - Gender",
         `Work absence statistics` %in% c("Days lost per worker in a year, illness or disability",
                                            "Days lost per worker in a year, personal or family responsibility"),
         REF_DATE >= 2015)
```

### Health care absence over time

```{r}
#| label: fig-healthcare-trend
#| fig-cap: "Days lost per worker per year, Health Care and Social Assistance, Canada, 2015–present"
#| echo: false

healthcare_absence %>%
  filter(`Work absence statistics` == "Total days lost per worker in a year",
         Gender == "Total - Gender") %>%
  ggplot(aes(x = REF_DATE, y = VALUE)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(x = "Year", y = "Days lost") +
  theme_minimal()
```

Health care workers lost 14 to 19 days per year over this period, with visible 
spikes in 2020 and 2022 followed by only partial recovery.

### How health care compares to the broader economy

```{r}
#| label: fig-healthcare-vs-service
#| fig-cap: "Health Care vs. All Service-Producing Industries"
#| echo: false

ggplot(compare_data, aes(x = REF_DATE, y = VALUE, 
                          color = `North American Industry Classification System (NAICS)`)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(x = "Year", y = "Days lost", color = "Industry") +
  theme_minimal()
```

Health care consistently ran 5 to 6 days higher per worker per year than the 
broader service-producing sector average across the entire period, a gap that 
persisted even as both series declined from their 2022 peak. This suggests the 
elevated absence burden in health care reflects a structural condition of the 
sector rather than a temporary pandemic effect.

### What's actually driving the gap: illness, not caregiving

```{r}
#| label: fig-healthcare-reason
#| fig-cap: "Health Care Absence, By Reason"
#| echo: false

ggplot(healthcare_split, aes(x = REF_DATE, y = VALUE, color = `Work absence statistics`)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(x = "Year", y = "Days lost") +
  theme_minimal()
```

Breaking the health care absence figure down by cause shows the gap is driven 
almost entirely by illness or disability, which ran 12 to 16+ days across the 
period, while personal or family responsibility stayed flat at 3 to 5 days 
throughout, barely moving even through 2020–2022.

### Takeaway

Health care and social assistance workers carry a persistently higher absence 
burden than the broader Canadian workforce, and that burden is driven by illness 
or disability rather than caregiving responsibilities. This points toward the 
occupational health cost of the work itself, high emotional demand, understaffing, 
and sustained patient-facing stress, rather than a work-life balance issue. It 
connects functional presence at work (simply showing up) to a separate, harder 
question: whether workers in high-demand roles are genuinely well, or working 
through conditions that eventually surface as time away from work entirely.
